---
title: "Stream Flow-Area Under Curves"
author: "Ephraim Romesberg"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(readxl)
library(ggplot2)
```


Loading the data:
```{r}
PL_Inflow <- read_excel("Data/1993-94 Water year PL Inflow.xlsx")
PL_Inflow$Date=as.Date(PL_Inflow$Date)
#converting the date to a numeric value (number of days from minimum date in data)
PL_Inflow$days_from_min=as.numeric(PL_Inflow$Date-min(PL_Inflow$Date))
```


Checking for incorrectly ordered and repeated values:
```{r}
PL_Inflow%>%group_by(Stream)%>%summarise(all(sort(days_from_min)==days_from_min))
PL_Inflow%>%group_by(Stream)%>%summarise(n_distinct(days_from_min)==length(days_from_min))
```

Taking the integral of the cubic spline interpolation of the sampled data points (the log/exp transformation is done to avoid getting negative y values):
```{r}
spline_int=function(x,y,method="natural",log_trans=FALSE){
  if (any(sort(x)!=x)) {stop("x values must be sorted in ascending order")}
  m=min(x)
  M=max(x)
  if (log_trans){
    s_f_l=splinefun(x,log(y),method = method)
    s_f=function(x){exp(s_f_l(x))}
  }
  else {
    s_f=splinefun(x,y,method = method)}
  int=integrate(s_f,lower = m,upper = M)$value
  average_val=int/(M-m)
  return(data.frame(Area_Under_Curve=int,Average_Flow_Rate=average_val))
}
```


Integration using the trapezoid rule:
```{r}
trapezoid_int=function(x,y){
  if (any(sort(x)!=x)) {stop("x values must be sorted in ascending order")}
  n=length(y)
  m=min(x)
  M=max(x)
  dx=diff(x)
  int=sum(((y[2:n]+y[1:(n-1)])/2)*dx)
  average_val=int/(M-m)
  return(data.frame(Area_Under_Curve=int,Average_Flow_Rate=average_val))
}
```


Calculating the integral and average function value using CS interpolation:
```{r}
average_flow_rates_spline=PL_Inflow%>%group_by(Stream)%>%summarise(spline_int(days_from_min,`Flow Rate (cfs)`,log_trans = TRUE))
average_flow_rates_spline$Area_Under_Curve=round(average_flow_rates_spline$Area_Under_Curve,2)
average_flow_rates_spline$Average_Flow_Rate=round(average_flow_rates_spline$Average_Flow_Rate,2)
average_flow_rates_spline
```

Using the trapezoid rule:
```{r}
average_flow_rates_trap=PL_Inflow%>%group_by(Stream)%>%summarise(trapezoid_int(days_from_min,`Flow Rate (cfs)`))
average_flow_rates_trap$Area_Under_Curve=round(average_flow_rates_trap$Area_Under_Curve,2)
average_flow_rates_trap$Average_Flow_Rate=round(average_flow_rates_trap$Average_Flow_Rate,2)
average_flow_rates_trap
```


Plotting the CS interpolation function for each stream:
```{r}
sf=function(stream_name,n=500,method="natural",log_trans=FALSE){
  data=filter(PL_Inflow,Stream==stream_name)
  x=data$days_from_min
  y=data$`Flow Rate (cfs)`
  if (log_trans){
    s_vals=spline(x,log(y),n=n,method=method)
    s_vals$y=exp(s_vals$y)
  }
  else {
    s_vals=spline(x,y,n=n,method=method)
  }
  return(data.frame(x=s_vals$x,y=s_vals$y))}
```


```{r}
stream_names=unique(PL_Inflow$Stream)
Spline_Functions=lapply(stream_names,FUN = sf,log_trans=TRUE)
names(Spline_Functions)=stream_names
Spline_Functions=bind_rows(Spline_Functions,.id = "Stream")
```


```{r}
avg_flow_rate_spline=paste(average_flow_rates_spline$Stream,paste(average_flow_rates_spline$Average_Flow_Rate, "cfs"),sep = ": ")
names(avg_flow_rate_spline)=average_flow_rates_spline$Stream
ggplot(data = Spline_Functions,aes(x,y,group = Stream))+geom_line(aes(colour = Stream))+geom_point(data = PL_Inflow,aes(x=days_from_min,y=`Flow Rate (cfs)`,group = Stream))+facet_wrap(~Stream,scales = "free_y",labeller =labeller(Stream=avg_flow_rate_spline))+labs(title = "Averages using Cubic Spline Interpolation",x="day",y="flow rate (cfs)")+theme_bw()
```

Plotting the linear interpolation function for each stream:
```{r}
avg_flow_rate_trap=paste(average_flow_rates_trap$Stream,paste(average_flow_rates_trap$Average_Flow_Rate, "cfs"),sep = ": ")
names(avg_flow_rate_trap)=average_flow_rates_trap$Stream
ggplot(data = PL_Inflow,aes(x=days_from_min,y=`Flow Rate (cfs)`,group = Stream))+geom_line(aes(colour = Stream))+geom_point()+facet_wrap(~Stream,scales = "free_y",labeller = labeller(Stream=avg_flow_rate_trap))+labs(title = "Averages using Linear Interpolation",x="day",y="flow rate (cfs)")+theme_bw()
```


Exporting the area under the curve (cfs-days) and average flow rate (cfs) values for each stream (using linear and cubic spline interpolation to approximate the integrand):
```{r}
Average_Flow_Rates=bind_rows(list("Spline"=average_flow_rates_spline,"Linear"=average_flow_rates_trap),.id = "Interpolation Method")
names(Average_Flow_Rates)[3:4]=c("Area_Under_Curve (cfs-days)", "Average_Flow_Rate (cfs)")
writexl::write_xlsx(Average_Flow_Rates,path = "Data/Average_Flow_Rates.xlsx")
```