---
title: "Trend Analysis of Priest Lake Water Temperature Data"
author: "Ephraim Romesberg"
date: "2024-12-31"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(zoo)
library(dplyr)
SCA_trend_Data <- read.csv("Data/SCA_trend_Data.csv")
SCA_Database <- read.csv("Data/SCA_Database.csv")
Tau_CI=read.csv("Data/Tau_CI.csv")
TS_CI=read.csv("Data/Theil_Sen_CI.csv")
SCA_trend_Data$Month=factor(SCA_trend_Data$Month,levels = c("Jun","Jul","Aug","Sep"),ordered = TRUE)
```

## The Data

-Data consists of temperature measurements from Priest Lake
-Only Data from top meter is used
-Data is only used from the sites KALI, PLNO, and PLSO and months June, July, August, and September
-Other sites and months contained too little data or had too large of gaps between observations

```{r,warning=FALSE}
 SCA_trend_Data$Date=as.Date(as.yearmon(paste(SCA_trend_Data$Month,SCA_trend_Data$Year,sep="-"),'%b-%Y'))
ggplot(data = SCA_trend_Data,aes(x=Date,y=Temperature))+geom_point(aes(colour = Site,shape = Month),size=1)+labs(title="Priest Lake Temperature Data",x="Date",y="Temperature (°C)")+theme_bw()
```

## Is temperature increasing over time?

Two ways to test this:

1) Fit a parametric function to model the central tendency (i.e. mean or median) of the data over time:
  -Simplest example of this is linear Regression (i.e. modeling temperature as a linear function of time) 
  -If the regression line has a positive slope this is evidence of an increasing trend

2) Look at the pairs of observations:
  -For each pair of successive observations $y_i,y_j$ measured at times $i<j$, how many increase over time (i.e. $y_i<y_j$) and how many decrease or stay the same? 
  -If more of the pairs increase over time than decrease this is evidence of an increasing trend

## Can we use Linear Regression?

```{r}
 SCA_trend_Data$Date=as.Date(as.yearmon(paste(SCA_trend_Data$Month,SCA_trend_Data$Year,sep="-"),'%b-%Y'))
ggplot(data = SCA_trend_Data,aes(x=Date,y=Temperature))+geom_point(aes(colour = Site,shape = Month),size=1)+geom_smooth(formula = y~x,method = "lm",se=FALSE,linewidth=1)+labs(title="Priest Lake Temperature Data",x="Date",y="Temperature (°C)")+theme_bw()
```


## Problems With Using Linear Regression:

- Assumes data is normally distributed
- Assumes Relationship between x and y is linear 
- Doesn't account for the Seasonal and Regional nature of the data
- Doesn't account for spatial or temporal dependence between observations (i.e. correlation between $Y_i$ and $Y_j$ is assumed to be zero for $i\neq j$)

## The Second Approach:

-The approach that looks at pairs of successive observations is known as the Mann-Kendall Test 
-This test looks at all possible pairs of sample data $y_i,y_j$ where $i<j$ 
-For each pair we calculate $sign(y_j-y_i)$, where $sign(y_j-y_i)=1$ if $y_j>y_i$, $0$ if $y_j=y_i$ and $-1$ if $y_j<y_i$
-We take the average of these values 
-This average is a value between $-1$ and $1$ and is known as Kendall's Tau
-If Kendall's Tau is positive this means that more of the pairs increase over time than decrease

## The Mann-Kendall Test:

-Null Hypothesis: Observations are equally likely to increase or decrease over time (no trend)
-Alternative Hypothesis: Observations are more likely to increase than decrease over time (positive trend)
-The Mann-Kendall Test tells us if Kendall's Tau is different enough from zero to conclude there is a trend 
-If the null hypothesis is true then (the sample) Kendall's Tau is approximately normal (for large enough sample sizes) with an expected value of zero
-Hypothesis test is constructed using the approximate normality of Kendall's Tau
-The approximate normality of Kendall's Tau holds regardless of whether the data is normal
-Mann-Kendall test makes no assumption that the relationship between x and y is linear
-Only depends on the ranks of the observations

## Issues with the Mann-Kendall Test:

Seasonal/Regional effects:

-While the MK test can deal with non-normal and non-linear data it is not designed to deal with seasonal or regional effects
-This can lead to misleading results
-When observations are compared across seasons or regions seasonal or regional differences can be mistaken for trend

Spatial/Temporal dependence:

-MK test also assumes that observations are independent
-This is often an unrealistic assumption
-It is natural for observations close in space and time to have some type of dependence

## Dealing with Seasonal/Regional Effects:

-Fortunately there are seasonal and regional versions of the MK test
-Rather than looking at all pairs of sample data we can only look at pairs within the same season and region
-This is Known as a Regional-Seasonal Kendall test
-We only compare pairs within each Region-Season combination
-These pairwise comparisons are then averaged like before to calculate Kendall's Tau 
-A positive R-S Tau tells us that there is an increasing trend in one or more Region-Season combination


## Dealing with Spatial and Temporal Dependence:

-R-S Kendall test still assumes that the observations are independent
-Using a test on correlated observations that assumes independence will often lead to misleading results
-A correlated sample provides less information than an independent sample
-We need to adjust for this loss of information when performing our hypothesis test

Dependent Sample=Less Information

-There is a version of the Seasonal Kendall test that is adjusted for serial dependence over time 
-For technical details see Hirsch and Slack (1984)
-This test is easily extended to also account for spatial dependence in regional data

## Which test is best for our data?

-We have no reason to assume our data is normal or that if there is a trend it is linear
-Our data is from multiple seasons (June, July, August, and September) and Regions (KALI, PLNO, and PLSO)
-More than likely there is spatial and temporal dependence between observations

This makes a Regional-Seasonal Kendall test adjusted for Spatiotemporal dependence a better choice than a more conventional method such as linear regression.

## R-S Test Results:

-R-S Kendall test adjusted for Spatiotemporal dependence was performed using `rkt` package in R
-Here regions are sites (KALI, PLNO, and PLSO) and  seasons are months (June, July, August, and September)
-Based on this test we got a Tau value of approximately `.27` and p-value of approximately  `0.005`
-Using a standard significance level of `.05` we reject the null hypothesis and conclude that there is an increasing trend, later observations are likely to be larger than earlier ones

## Trend Within Seasons:

-The R-S Kendall test tells us that there is a trend in at least some of the region-season combinations 
-It does not tell us which specific regions and seasons have a trend
-It seems plausible that warmer months could have more of a trend than cooler months
-We can investigate this further by running individual Regional Kendall tests for data from each month

Results From By-Month tests:

-Greatest trend is in July (Tau value of `0.49` and p-value of `0.002`)
-Weakly significant positive trends in June and August
-No significant trend in September

## Quantifying the Trend:

-A positive Kendall's Tau tells us that observations are more likely to increase over time
-It does not tell us how much they increase (i.e. °C/Year)
-To estimate the magnitude of the trend we have another statistic known as the Theil-Sen's slope
-Theil-Sen's slope also uses pairs of successive observations
-To calculate Theil-Sen's slope we calculate the slope of the line connecting each pair of observations, we then take the median of these slopes
-This approach can be adjusted for regional/seasonal data just as with Kendall's Tau
-Theil-Sen slope does not assume normality and is much less sensitive to outliers than ordinary regression slope
-It does however assume the trend is linear, so interpret with caution 

## Theil-Sen Slopes:

-The Regional-Seasonal Theil-Sen Slope is approximately `.08`
-So (as a very rough approximation) we can say that Temperature is increasing at about .08 °C/Year
-Since some months have a more significant trend than others it is better to look at the slopes by month
-We get the largest slope in July at approximately 0.14 °C/Year

## Confidence Intervals for Tau:

```{r}
Tau_CI$season=factor(Tau_CI$season,levels = c("RS.Tau","Jun","Jul","Aug","Sep"),ordered = TRUE)
Tau_CI$P.Value=factor(c("<.01","Not Significant","<.01","<.05","Not Significant"),levels = c("<.01","<.05","Not Significant"))
ggplot(Tau_CI)+geom_crossbar(aes(x=season,y=tau,ymin = X2.5..,ymax = X97.5..,colour = P.Value),width=.25)+geom_hline(yintercept = 0)+labs(title = "Kendall's Tau Values with 95% Confidence Intervals")+theme_bw()
```

## Confidence Intervals for Theil-Sen's Slope:

```{r}
TS_CI$season=factor(TS_CI$season,levels = c("RS.Slope","Jun","Jul","Aug","Sep"),ordered = TRUE)
ggplot(TS_CI)+geom_crossbar(aes(x=season,y=TS.slope,ymin = X2.5..,ymax = X97.5..),width=.2)+geom_hline(yintercept = 0)+labs(title = "Theil-Sen's Slope Values with 95% Confidence Intervals")+theme_bw()
```

## Interpreting the Confidence Intervals:

-The confidence intervals give us a range of plausible values for Tau and TS-Slope values
-If a confidence interval contains zero this indicates that the trend is not significant
-We get the same results from the CI's as we did from the hypothesis tests
-Overall Regional Seasonal trend is positive 
-By season we have the greatest trend and largest slope in July

## Conclusion:

-The Regional-Seasonal Kendall Test relies on much less restrictive assumptions than linear regression
-Data does not need to be normal and relationship does not need to be linear
-Test accounts for the regional and seasonal nature of the data and
spatial/temporal dependence
-Based on this test we can conclude that the water temperature is increasing over
time in Priest Lake
-The most significant trend and largest rate of increase is in July

## Refrences:

Design Trend Seasonal Kendall. (n.d.). Vsp.pnnl.gov. https://vsp.pnnl.gov/help/Vsample/Design_Trend_Seasonal_Kendall.htm

Helsel, D. R., & Hirsch, R. M. (2020). Statistical methods in water resources (pp. 342–349). U.S. Geological Survey.

Hirsch, R. M., & Slack, J. R. (1984). A Nonparametric Trend Test for Seasonal Data With Serial Dependence. Water Resources Research, 20(6), 727–732. https://doi.org/10.1029/wr020i006p00727

Mann, H. B. (1945). Nonparametric Tests Against Trend. Econometrica, 13(3), 245. https://doi.org/10.2307/1907187‌

Marchetto, A. (2024). rkt: Mann-Kendall Test, Seasonal and Regional Kendall Tests. R-Project.org. https://cran.r-project.org/package=rkt




